---
title: "ASA Data expo 2006"
bibliography: ../refs.bib
output: html_document
---
```{r}
library(tidyverse)
data("nasa", package = "GGally")
```

In this example, we re-visit a data set that was used for the ASA Data Expo in 2006. The `nasa` data, made available as part of the `ggpcp` package provides an extension to the data provided in the `GGally` package [@GGally]. It consists of monthly measurements of several climate variables, such as cloud coverage, temperature, pressure, and ozone values, captured on a 24x24 grid across Central America between 1995 and 2000.

Using a hierarchical clustering (based on Ward's distance) of all January  measurements of all climate variables and the elevation, we group locations into 6 clusters.
The resulting cluster membership can then be summarized visually. @fig:spatial shows a tile plot of the geography colored by cluster. We see that the clusters have a very distinct geographic pattern.

```{r prep-work-nasa}
wide <- nasa %>% dplyr::select(id, month, year, starts_with("cloud"), ozone, pressure, surftemp, elevation) %>%
  gather(key = variable, value = value, -id, -month, -year) %>%
  filter(month %in% c("Jan")) %>%
  unite(var_date, variable, year, month, sep="-") %>%
  spread(key=var_date, value)

d <- dist(wide[,-1])
cl <- hclust(d, method = "ward.D")

wide$cl2 <- cutree(cl, k = 2)
wide$cl3 <- cutree(cl, k = 3)
wide$cl4 <- cutree(cl, k = 4)
wide$cl5 <- cutree(cl, k = 5)
wide$cl6 <- cutree(cl, k = 6)
wide$cl7 <- cutree(cl, k = 7)
wide$cl8 <- cutree(cl, k = 8)
wide$cl9 <- cutree(cl, k = 9)
wide$cl10 <- cutree(cl, k = 10)

wide.sum <- nasa %>% dplyr::select(id, month, year, starts_with("cloud"), ozone, pressure, surftemp, elevation) %>%
  gather(key = variable, value = value, -id, -month, -year) %>%
  filter(month %in% c("Jan")) %>%
  group_by(id, variable, month) %>%
  summarize(
    value = mean(value)
  ) %>%
  unite(var_month, variable, month, sep="-") %>%
  spread(var_month, value)

wide <- wide %>% left_join(wide.sum, by="id")
```


```{r spatial-cropped, fig.align="center", out.width="\\linewidth", fig.cap="(Left) Tile plot of the (gridded) geographic area underlying the data. Each tile is colored by its cluster membership. (Right) Generalized Parallel Coordinate Plot of the cluster means for atmospheric variables."}
knitr::include_graphics("figures/spatial-1-cropped.pdf")
```


```{r spatial, fig.align="center", eval=FALSE, out.width="\\linewidth", fig.cap="Tile plot of the (gridded) geographic area underlying the data. Each tile is colored by its cluster membership."}
names(wide)[56] <- "elevation"
means <- wide %>%
  group_by(cl8) %>%
  summarize(across(`cloudhigh-Jan`:`surftemp-Jan`, .fns = mean))
means$type <- "cluster-means"

nasa_pcp <- wide %>%
  mutate(
    type = "observed") %>%
  bind_rows(means) %>%
  pcp_select(`cloudhigh-Jan`:`surftemp-Jan`) %>%
  pcp_scale()


gg1 <- wide %>% separate(id, into=c("y", "x"), remove = FALSE) %>%
  mutate(x = as.numeric(x), y = as.numeric(y)) %>%
  ggplot(aes(x = x, y=y, fill=factor(cl8))) +
  geom_tile(alpha = 0.7, colour = "grey90", size=0.1) + scale_fill_brewer("Cluster", palette = "Paired") +
  xlab("Latitude") + ylab("Longitude") +
  coord_equal() +
  ggthemes::theme_map() +
  theme(legend.position = "none") +
     theme(panel.grid = element_blank(),
        #   panel.background = element_rect(fill = "#FFFFFF"),
       #    plot.background = element_rect(fill = "#FFFFFF"),
           plot.margin = rep(unit(0,"null"),4),
           panel.spacing = unit(0,"null"),
           panel.border     = element_blank())



gg2 <- nasa_pcp %>%
  filter(type == "cluster-means") %>%
  ggplot(aes_pcp()) +
  geom_pcp(aes(colour=factor(cl8)), size = 1.25) +
  scale_colour_brewer("Cluster", palette = "Paired") +
  theme_bw() +
    theme(legend.position = "bottom") +
  ylab("") + xlab("") +
  guides(colour=guide_legend(override.aes = list(alpha=1, size = 2))) +
  theme(axis.text.x = element_text(angle = 60, vjust = 0.5, hjust=0.5)) +
  coord_equal()


gg1+gg2 + plot_layout(widths=c(1,2))
```


```{r clusters, fig.height = 5, fig.width = 8, fig.pos="H", fig.cap="Overview of all variables involved in the clustering. ", out.width='\\textwidth'}
nasa_pcp %>%
  filter(type == "observed") %>%
  ggplot(aes_pcp()) +
  geom_pcp(aes(colour=factor(cl8)), alpha = 0.3) +
  geom_pcp(#aes(colour=factor(cl8)),
    size=1, alpha = 0.6,
           data = nasa_pcp %>% filter(type == "cluster-means")) +
  facet_wrap(.~cl8, labeller="label_both", ncol=2) +
  scale_colour_brewer("Cluster", palette = "Paired") +
  theme(legend.position = "none") +
  theme_bw() +
  ylab("") + xlab("") +
  guides(colour=guide_legend(override.aes = list(alpha=1, size = 2)))
```

\hh{XXX changed the data for the clustering - need to adjust the description}
From the parallel coordinate plot in @fig:clusters we see that cloud coverage in low, medium and high altitude distinguishes quite succinctly between some of the clusters. (Relative) temperatures in January are very effective at separating between clusters in the Southern and Northern hemisphere.
The connection between the US gulf coast line and the upper region of the Amazon (cluster 2) can probably be explained by a relatively low elevation combined with similar humidity levels.

A parallel coordinate plot allows us to visualize a part of the dendrogram corresponding to the hierarchical clustering.

```{r old,fig.height = 3.5, out.width='.85\\linewidth', fig.cap="Dendrogram showing number of clusters at each step of the hierarchical process using the old-style parallel coordinate plot."}
wide %>%
  pcp_select(cl2:cl10) %>%
  pcp_scale %>%
  ggplot(aes_pcp()) +
  geom_pcp() +
  theme_bw() +
  xlab("Number of clusters") +
  ylab("")
```

Using the generalized parallel coordinate plots we can visualize the clustering process in plots similar to what @clustergram1, @clustergram2 coined the clustergram, see @fig:old and @fig:clustergram.

```{r clustergram, fig.height = 3.5, fig.cap="Same dendrogram as above using the much more informative generalized parallel coordinate plot."}
wide %>%
  mutate(
    cl2 = factor(cl2),
    cl3 = factor(cl3),
    cl4 = factor(cl4),
    cl5 = factor(cl5),
    cl6 = factor(cl6),
    cl7 = factor(cl7),
    cl8 = factor(cl8),
    cl9 = factor(cl9),
    cl10 = factor(cl10)
  ) %>%
  pcp_select(cl2:cl10) %>%
  pcp_scale() %>%
  pcp_arrange() %>%
  ggplot(aes_pcp()) +
  geom_pcp(alpha = 0.5, aes(colour = factor(cl10))) +
  geom_pcp_boxes( boxwidth=0.1, fill="white", alpha = 0.3) +
  xlab("Number of clusters") +
  scale_colour_brewer("Cluster", palette="Paired") +
  ylab("") +
  guides(colour=guide_legend(override.aes = list(alpha=1)))
```

Along the x-axis the number of clusters are plotted with one PCP axis each, from two clusters (left) to 10 clusters (right most PCP axis). Each line corresponds to one location, lines are colored by cluster assignment in the ten-cluster solution. This essentially replicates  the dendrogram while providing information about the number of observations in each cluster as well as the relationship between successive clustering steps.
